---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hydromad

<!-- badges: start -->
<!-- badges: end -->

The goal of hydromad is to provide a modelling framework for environmental hydrology: water balance accounting and flow routing in spatially aggregated catchments. 
  
Hydromad supports simulation, estimation, assessment and visualisation of flow response to time series of rainfall and other drivers. A minimal unit hydrograph framework is used, where areal rainfall is passed through a soil moisture accounting (SMA) model to estimate effective rainfall; this is then passed through a routing model to estimate streamflow. Included are several implementations of common hydrological models consistent with this framework.

## Installation

<!--- You can install the released version of hydromad from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("hydromad")
```
--->
Currently you can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JosephGuillaume/hydromad")
```
## Usage

This is a basic example which shows you how to solve a common problem:

```{r example, message = F}
library(hydromad)
## basic example code
data(Cotter)
## IHACRES CWI model with exponential unit hydrograph
## an unfitted model, with ranges of possible parameter values
modx <- hydromad(Cotter[1:1000], sma = "cwi", routing = "expuh",
                 tau_s = c(2,100), v_s = c(0,1))
modx
```
