---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hydromad

<!-- badges: start -->
[![R build status](https://github.com/josephguillaume/hydromad/workflows/R-CMD-check/badge.svg)](https://github.com/josephguillaume/hydromad/actions)
[![codecov](https://codecov.io/gh/josephguillaume/hydromad/branch/master/graph/badge.svg)](https://codecov.io/gh/josephguillaume/hydromad)
<!-- badges: end -->

The goal of hydromad is to provide a modelling framework for environmental hydrology: water balance accounting and flow routing in spatially aggregated catchments. 
  
Hydromad supports simulation, estimation, assessment and visualisation of flow response to time series of rainfall and other drivers. A minimal unit hydrograph framework is used, where areal rainfall is passed through a soil moisture accounting (SMA) model to estimate effective rainfall; this is then passed through a routing model to estimate streamflow. Included are several implementations of common hydrological models consistent with this framework.

The main resource for hydromad is the website: [http://hydromad.catchment.org](http://hydromad.catchment.org)
The website describes the different functions and gives examples of the code applications.

## Installation

<!--- You can install the released version of hydromad from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("hydromad")
```
--->
* Add the `hydromad` repository for this R session:
```r
options(repos=c("http://hydromad.catchment.org",getOption("repos")))
```
* Install hydromad:
``` r
install.packages("hydromad")
``` 

* Optionally install other packages required for certain functions:
``` r
install.packages("DEoptim")
install.packages("dream")
``` 

Once it is installed, get started with

``` r   
library(package = "hydromad")
help("hydromad")
```  

The development version can be installed from [GitHub](https://github.com/) with:

``` r  
# install.packages("devtools")
devtools::install_github("JosephGuillaume/hydromad")
```

## Contributing

The maintainers of hydromad are always keen to receive contributions. However to help us we would like you to consider the following:

* Please read and use our [CONTRIBUTING document](https://github.com/josephguillaume/hydromad/blob/master/docs/CONTRIBUTING.md), this will really help us integrating your solution into hydromad.  
* This is a list of the [current issues](https://github.com/josephguillaume/hydromad/issues) (which you might be able to help with).  
* As part of contributing we would like you to consider the ropensci [Code of Conduct](https://ropensci.org/code-of-conduct/), which we use for hydromad. 


## Usage

This is a basic example of defining a model in hydromad:

```{r example, message = F}
library(hydromad)
## basic example code
data(Cotter)
## IHACRES CWI model with exponential unit hydrograph
## an unfitted model, with ranges of possible parameter values
modx <- hydromad(Cotter[1:1000], sma = "cwi", routing = "expuh",
                 tau_s = c(2, 100), v_s = c(0, 1))
modx
```

## License

This project is licensed under the terms of [GPL (>=2)](https://github.com/josephguillaume/hydromad/blob/master/DESCRIPTION)


## Acknowledgements

This research was supported by the Sydney Informatics Hub, a Core Research Facility of the University of Sydney.
